# Tutorial 3: Data Visualisation for Business Intelligence {.unnumbered}

::: {.callout-tip}
## Learning Goals {.unnumbered}

By the end of this tutorial, you should be able to:

* Explain how different `ggplot2` visualization techniques reveal patterns in the Melbourne housing market that would be difficult to see in tables or raw numbers.
* Use `ggplot2` to create histograms, and scatterplots that analyze relationships between property prices, location, and dwelling types.
* Apply transformations such as log scales, proportions, and faceting to improve the interpretability of price distributions and spatial trends.
* Customize `ggplot2` plots using themes, color palettes, and annotations to enhance clarity and storytelling.
* Evaluate and compare visualization approaches, identifying which effectively communicate insights about housing price trends and spatial patterns.
* Build and refine a visual analysis workflow that could inform practical decision-making for home buyers, investors, or urban planners
:::

## The Business Challenge

```{r}
#| warning: false
#| echo: false
library(tidyverse)  # for plotting, includes ggplot
library(patchwork)  # for combining multiple plots into subfigures
library(scales)     # for formatting axis scales
library(ggokabeito) # color blind friendly color palette -- this course's default
```

```{r}
#| warning: false
#| eval: true
#| echo: false
#| cache: true

# Create data/ directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}
# Define URL and destination file path
url <- "https://github.com/unimelb-cmce-10002/fba-book/raw/refs/heads/main/data/melbourne_housing.csv"
destfile <- "data/melbourne_housing.csv"
# Download the file if it doesn't already exist
if (!file.exists(destfile)) {
  download.file(url, destfile, mode = "wb")
}
# Load the data
housing <- read_csv(destfile)
```

### The Topic: Understanding Melbourne’s Housing Market {.unnumbered}

Melbourne’s real estate market is one of the most dynamic in Australia. Property prices are influenced by several factors, including location, dwelling type, and proximity to the Central Business District (CBD). Buyers, sellers, and policymakers constantly analyze housing trends to make informed decisions. Understanding these trends is crucial in a $2 trillion+ housing market, where even small fluctuations in pricing expectations can have significant financial impacts.

A key question we will explore is: How do property prices vary by region, dwelling type, and location? By analyzing real estate data, we can identify patterns that help explain the drivers of Melbourne’s housing prices and their implications for different stakeholders in the market.

### The Data: The Melbourne Housing Market Dataset {.unnumbered}

We will use a dataset containing real estate sales data from Melbourne. The dataset provides insights into housing trends and includes key variables that influence property prices.

The dataset captures the property location within broader metropolitan regions (regionname), allowing us to compare trends across different areas. It also includes the type of dwelling, classified as House (h), Townhouse (t), or Apartment (u), enabling us to assess price differences based on housing type. The sale price (price) provides a measure of market value, while the distance to the CBD (distance) helps analyze the impact of proximity to the city center on property prices.

Similar datasets are widely used by real estate analysts, banks, and property investors to assess housing trends, predict future movements, and guide investment decisions. Through this tutorial, we’ll work with real-world data to uncover patterns and develop a deeper understanding of Melbourne’s property market. Let's dive in!

### Where we're headed {.unnumbered}

Just a few lines of R code transform numbers to data visualizations.

From this:

```{r}
#| echo: false 
housing %>%
    select(type, address, postcode, price, distance, regionname) %>%
    head(10)
```

to this:

```{r}
#| fig-width: 7
#| fig-height: 7
#| fig-align: "center"
#| echo: false

price_hist <- 
    ggplot(housing,
           aes(x = price, fill = type)
           ) +
    geom_histogram(
        bins = 50
    ) +
    # Log-scale the x-axis to handle price skew
    scale_x_continuous(
        trans = "log10",
        labels = scales::dollar
    ) +
    # Add labels and title
    labs(
        x = "Price", 
        y = "Number of Listings", 
        title = "Prices by Dwelling Type"
    ) +
    # Create separate plots for each dwelling type
    facet_wrap(~type) +
    # remove grey background
    theme_bw() +
    # Rotate x-axis labels for readability
    theme(axis.text.x = element_text(angle = 55, hjust = 1)) +
    # Use colorblind-friendly fill palette
    scale_fill_okabe_ito() +
    # Center the title
    theme(
        plot.title = element_text(hjust = 0.5)
    )

scttr <-
    ggplot(housing,
           aes(x = distance, y = price)
           ) +
     # Scatter plot with reduced alpha to handle overplotting
    geom_point(alpha = 0.05, color = "grey") +
    # Polynomial fit to capture curvature (not just a straight line)
    geom_smooth(
        method = "lm",
        se = FALSE,
        color = "coral",
        linewidth = 1
    ) +
    # Format y-axis as dollars
    scale_y_continuous(labels = scales::dollar) +
    # Add labels and center the title
    labs(
        x = "Distance to CBD (km)",
        y = "Price",
        title = "Dwelling Price vs. Distance"
    ) +
    theme(
        plot.title = element_text(hjust = 0.5)
    ) + 
    theme_bw() +
    theme(legend.position = "none")

region_bar <- 
    # 2. Create the plot
    ggplot(housing,
        # 2a. Reorder the region names based on frequency
        aes(x = fct_infreq(regionname))
    ) +
    # 3. Plot bar heights as proportions, not raw counts
    geom_bar(
      # THIS BIT IS HARD
      # YOU CAN NOT USE IT AND KEEP Y AXIS AS COUNTS
        aes(y = after_stat(count) / sum(after_stat(count))),
        fill = "grey"
    ) +
    # 4. Add informative axis labels and a title
    labs(
        x = "Region", 
        y = "Frequency",  # Note: this is technically a proportion, not % format
        title = "Number of Listings by Region"
    ) +
  # 5. Tweak the theme for better readability and presentation
    scale_y_continuous(labels = scales::percent)  + 
    theme_bw() +
    theme(
        axis.text.x = element_text(angle = 25, hjust = 1),  # Rotate x labels
        plot.title = element_text(hjust = 0.5)              # Center the title
    )

# Bringing together 
price_hist /    (region_bar | scttr)  & theme(legend.position = "none")

```

## Loading the Data

### R packages for today {.unnumbered}

```{r}
#| eval: false
library(tidyverse)  # for plotting, includes ggplot
library(patchwork)  # for combining multiple plots into subfigures
library(scales)     # for formatting axis scales
library(ggokabeito) # color blind friendly color palette -- this course's default
```

### Loading the Data in R {.unnumbered}

```{r}
#| warning: false
#| eval: false
#| echo: true
#| label: load-housing-data
#| cache: true

housing <-
    read_csv("data/melbourne_housing.csv")
```

## Prepare these Exercises before Class 

Prepare these exercises before coming to class. Plan to spend 45 minutes on these exercises.

::: {.callout-important title = "Switch the `eval` flag to TRUE when you want to evaluate code!"}
In the R code chunks below we have provided starter code for you to work from.
We have set the key `eval` to the value `false` so that they are not run because they have syntax such as `YOUR_VALUE_HERE` which would generate errors.

Switch the `eval` value to `TRUE` when you want the R code within a chunk to be run when you compile your document.
:::

### Exercise 1: Understanding `ggplot2` Syntax

Consider the code below:

```{r}
#| eval: false

ggplot(housing, 
       aes(x = regionname)
       ) +
  geom_bar() +
  labs(
    x = "Region",
    y = "Count",
    title = "Number of Listings by Region"
  )
```

**(a)** What does `ggplot(housing, aes(x = regionname))` do?

**(b)** What is the role of `aes(x = regionname)` in this code?

**(c)** What does `geom_bar()` add to the plot?

**(d)** What does `labs()` do?

**(e)** In this code, what does the `+` operator do?

**(f)** Run the code above to produce the bar plot.

### Exercise 2: Visualizing the distribution of Prices by Dwelling Type

By the end of this exercise, your plot should look similar to this one:

```{r}
#| fig-width: 5
#| fig-height: 5
#| fig-align: "center"
#| echo: false

housing |>
    ggplot(aes(x = price)
           ) +
    geom_histogram(bins = 50) +
    labs(x = "Price", 
         y = "Density within Type", 
         title = "Prices by Dwelling Type"
         ) + 
    facet_wrap(~ type)
```

**(a)** Why is a histogram useful for understanding price distributions? What insights can we gain from plotting price distributions separately for each dwelling type?

:::{.callout-tip title="What is a Histogram?" collapse=true}
A **histogram** is a visualization that represents the **distribution of numerical data** by dividing values into bins and counting the number of observations in each bin. Unlike a bar plot, which displays **categorical data**, a histogram is used for **continuous variables** like price, age, or distance.

Histograms help identify **patterns** in data, such as **skewness, outliers, and central tendencies**, making them a crucial tool for understanding how values are distributed in a dataset.
:::

**(b)** Use the starter code below to create a basic histogram of house prices. Replace `YOUR_VARIABLE_NAME` with the correct variable.

```{r}
#| eval: false
ggplot(housing, 
       aes(x = YOUR_VARIABLE_NAME)
       ) +
    geom_histogram(bins = 50)
```

**(c)** Extend the plot above by adding labels to the x-axis, y-axis, and title using the `labs()` function.

```{r}
#| eval: false
ggplot(housing, 
       aes(x = YOUR_VARIABLE_NAME)
       ) +
    geom_histogram(bins = 50) +
    labs(x = "YOUR_X_AXIS_LABEL", 
         y = "YOUR_Y_AXIS_LABEL", 
         title = "YOUR_TITLE"
         )
```

**(d)** Modify the plot so that you create separate histograms for each dwelling type by adding the `facet_wrap(~type)` layer to the plot

```{r}
#| eval: false

ggplot(housing, 
       aes(x = price)
       ) +
    geom_histogram(bins = 50) +
    labs(x = "YOUR_X_AXIS_LABEL", 
         y = "YOUR_Y_AXIS_LABEL", 
         title = "YOUR_TITLE"
         ) 
    YOUR_CODE_HERE
```

**(e)** What does the histogram tell you about price differences between dwelling types?

### Exercise 3: The Price - Distance Relationship

By the end of this exercise, your plot should look similar to this one:

```{r}
#| fig-width: 5
#| fig-height: 5
#| fig-align: "center"
#| echo: false
ggplot(housing, 
       aes(x = distance, y = price)
       ) +
    geom_point(alpha = 0.05) +
    geom_smooth(method = "lm",
                se = FALSE, 
                linewidth = 1
                ) +
    labs(x = "Distance to CBD (km)", 
         y = "Price", 
         title = "Dwelling Price vs. Distance"
         )
```
**(a)** Why might we expect a relationship between price and distance to the CBD?

**(b)** Use the starter code below to build a scatter plot of **price vs. distance**, replacing `YOUR_VARIABLE_NAME` with the correct column names.

::: {.callout-tip title="What is a Scatter Plot?" collapse=true}
A **scatter plot** is used to visualize the relationship between two continuous variables by plotting individual data points. Each dot represents one observation, helping identify **patterns, clusters, and trends**. In this case, it allows us to explore how **house prices change with distance**.
:::

```{r}
#| eval: false
ggplot(housing, 
       aes(x = YOUR_VARIABLE_NAME, y = YOUR_VARIABLE_NAME)
       ) +
    geom_point()
```

**(c)** Since we have a large number of data points, many overlap, making it difficult to distinguish individual observations. Modify the `geom_point()` function to reduce over-plotting by adjusting the `alpha` value. Experiment with different values (e.g., 0.1, 0.05, 0.01) and describe how each affects the visualization.

```{r}
#| eval: false
ggplot(housing, 
       aes(x = YOUR_VARIABLE_NAME, y = YOUR_VARIABLE_NAME)
       ) +
    geom_point(alpha = YOUR_NUMBER)
```

**(d)** We can better understand the relationship by adding a statistical transformation to the plot. Add a fitted line using `geom_smooth(method = "lm", se = FALSE)`.

```{r}
#| eval: false
ggplot(housing, 
       aes(x = YOUR_VARIABLE_NAME, 
           y = YOUR_VARIABLE_NAME)
       ) +
    geom_point(alpha = YOUR_NUMBER) +
    YOUR_CODE_HERE
```

**(e)** Finally, add labels to the x-axis, y-axis, and title using the `labs()` function. Adjust the x- and y-axis scales as you deem appropriate.

```{r}
#| eval: false
ggplot(housing, 
       aes(x = YOUR_VARIABLE_NAME, y = YOUR_VARIABLE_NAME)
       ) +
    geom_point(alpha = YOUR_NUMBER) +
    YOUR_CODE_HERE +
    labs(x = "YOUR_LABEL", 
         y = "YOUR_LABEL", 
         title = "YOUR_TITLE"
         )
```

**(f)** What does the scatter plot reveal about the relationship between house prices and distance to the CBD? How does the line help interpret this relationship?

## In-Class Exercises

You will discuss these exercises in class with your peers in small groups and with your tutor. These exercises build from the exercises you have prepared above, you will get the most value from the class if you have completed those above before coming to class.

### Exercise 4: Improving the Price Distribution Plot

**(a).** Together with your peers, propose three changes you want to make to the plot in Exercise 2. Discuss the rationale behind each change and how it might improve the visualization.

**(b).** Work with your tutor to create an agreed upon list of changes to make to the plot. What steps do you need to take to make these improvements?

**(c).** Implement the changes suggested in the R code.


### Exercise 5: Improving the Price-Distance Relationship Plot

**(a).** Together with your peers, propose two changes you want to make to the plot in Exercise 3. Discuss the rationale behind each change and how it might improve the visualization.

**(b).** Work with your tutor to create an agreed upon list of changes to make to the plot. What steps do you need to take to make these improvements?

**(c).** Implement the changes suggested in the R.


### Exercise 6 [Optional, if time]: Improving the Listings by Region Plot

**(a).** Together with your peers, propose two changes you want to make to the plot in Exercise 1. Discuss the rationale behind each change and how it might improve the visualization.

**(b).** Work with your tutor to create an agreed upon list of changes to make to the plot. What steps do you need to take to make these improvements?

**(c).** Implement the changes suggested in the R.


### Exercise 7: Putting the plots together

**(a).** Use the code below to combine the plots from the exercises above. Replace `PLOT_ONE`, `PLOT_TWO`, and `PLOT_THREE` with the plots you created in Exercises 4, 5, and 6.

```{r}
#| eval: false
PLOT_ONE /    
    (PLOT_TWO | PLOT_THREE) & 
    theme(legend.position = "none")
```

**(b).** Save this combined plot as a PDF file using the `ggsave()` function. You can specify the file name and dimensions using the `filename` and `width` and `height` arguments.

```{r}
#| eval: false
ggsave("YOUR_FILE_NAME", width = 12, height = 8)
```


### Exercise 8: Synthesizing the Findings

If you were presenting these results in a business meeting, how would you explain the key takeaways? Structure your summary to include:

* What you analyzed (data, variables)
* What you found (key insights from the visualizations)
* What decisions could be made based on this information?

