---
title: 'Tutorial 10: Storing & Retrieving Data'
author: Foundations of Business Analytics
date: today
date-format: "MMMM, YYYY"
---

::: {.callout-tip}
## Learning Goals {.unnumbered}

By the end of this tutorial, you should be able to:

* Apply tidyverse workflows across alternative data sources (CSV, JSON Lines, DuckDB) to wrangle, flatten, and join auction data into tidy tables suitable for analysis.

* Analyse outcomes and telemetry by creating visuals and simple metrics (`seller_share`, `win_rate`) to identify patterns that could indicate shill behaviour.
:::

## The Business Challenge: Identifying Shill Bidding

Online auctions are big business. Platforms like eBay, Yahoo! Auctions, and specialist marketplaces connect millions of buyers and sellers every day. But with opportunity comes risk: not every bidder is playing fair.

One form of fraud is called **shill bidding**. This happens when sellers (or their associates) secretly bid on their own items. The goal isn't to win the auction, but to drive up the price and trick genuine buyers into paying more than they otherwise would. Shill bidders might:

- Place bids just high enough to outbid real customers, forcing them to keep bidding.
- Use multiple accounts or devices to hide their identity.
- Bid early and often to create the illusion of demand.

For an auction platform, shill bidding creates serious problems:

- It damages buyer trust, which can reduce participation and long-term revenue.
- It raises legal and reputational risks if fraud is not detected.
- It can distort the competitive landscape for honest sellers.

Your task in this exercise is to step into the role of an analyst at an online auction company. You've been asked to examine bidding data and look for patterns that might indicate suspicious activity. To do this, you'll need to:

1. Explore the auctions data provided in multiple formats (CSV, JSON, database).
2. Use wrangling and visualisation tools to spot differences between ordinary buyers and suspicious bidders.
3. Interpret these patterns in business terms: would a manager at the auction platform be worried?

## The Data: Auction Bids and Devices

We'll be working with synthetic data that simulates an online auction platform. It comes in a few different formats:

- CSV / DuckDB: structured data on auctions, bidders, bid amounts, and timestamps.
- JSON: extra details about each bidder's activity, such as the devices or sessions they used.

Together, these files let us see not just who is bidding and how much, but also how they are bidding, which can be a clue for spotting suspicious behaviour.

## The Method: Wrangling Data Across Backends

In this tutorial, the focus isn't just on wrangling data. It is on seeing how the same tidyverse tools work across different backends. We'll be using:

- `dplyr`: to filter, group, summarise, and join data.
- `tidyr`: to reshape nested or messy data into tidy tables.

You'll apply these verbs on top of different data sources:

- Flat files (CSV)
- Databases (DuckDB)
- Nested structures (JSON)

The key idea is that once you know the tidyverse workflow, it carries over no matter where the data lives.

```{r}
#| message: false
library(tidyverse) # Core tidyverse for wrangling and plotting
library(patchwork) # Combine multiple plots
library(jsonlite) # For reading JSON files
library(DBI) # For working with databases
library(duckdb) # for working with duckdb databases
```

## Prepare these Exercises before Class

Prepare these exercises before coming to class. Plan to spend 30 minutes on these exercises.

### Exercise 1: Auction Results Baseline

**(a)**. Read the CSV file located at `data/auction_results.csv` and examine its structure by completing the code below:

```{r}
#| eval: false
# Load the data
auction_results <-
    YOUR_CODE("data/auction_results.csv")

# Examine the structure
YOUR_CODE(auction_results)
```

**(b)**. What does one row represent in the dataset? Which columns look most informative for understanding auction outcomes?

**(c)**. Produce 3 plots using the starter code below:

1. A histogram of `final_price`
2. A histogram of `num_bids`
3. A scatterplot of the relationship between `final_price` and `num_bids`.

```{r}
#| eval: false

p1 <-
    auction_results |>
    ggplot() +
    geom_histogram(aes(x = YOUR_CODE)) +
    theme_minimal()

p2 <-
    auction_results |>
    ggplot() +
    YOUR_CODE(aes(x = YOUR_CODE), bins = YOUR_CODE) +
    theme_minimal()

p3 <-
    auction_results |>
    ggplot(aes(x = YOUR_CODE, y = YOUR_CODE)) +
    YOUR_CODE() +
    geom_smooth(method = "lm") +
    theme_minimal()

p1 / p2 / p3
```

**(d)**. Provide a short explanation of your findings from (c). Are there any patterns that are suggestive of shill bidding?

**(e)**. Next, let's explore how certain auction outcomes vary across sellers. Complete the starter code below to produce summary statistics by seller:

```{r}
#| eval: false
seller_profile <- 
    auction_results |>
    group_by(seller_id) |>
    summarise(
        n_auctions    = YOUR_CODE(),
        median_price  = YOUR_CODE(final_price, na.rm = TRUE),
        median_bids   = YOUR_CODE(num_bids, na.rm = TRUE),
        share_missing_winner = YOUR_CODE(is.na(YOUR_CODE)) / n_auctions,
        .groups = "drop"
    ) |>
    arrange(YOUR_CODE(n_auctions))
```

**(f)**. Based on the seller profiles in (e), are there any patterns suggestive of shill bidding? What further information would you need to confirm any suspicion?

### Exercise 2: Digging Deeper into User Behaviour

In Exercise 1, we built a market baseline from auction outcomes: prices, bid counts, and seller summaries. That showed us where to look, but not how those outcomes happened.
Next, we'll bring in behavioural telemetry from a JSON file (devices/sessions). Using `dplyr`/`tidyr` verbs, we'll flatten the nested JSON and link it to our auctions so we can ask: who was active, on which devices, and how activity clustered around particular sellers.

**(a)**. An example record in the JSON file `data/bids.jsonl` is as follows: 

```json
{
  "auction_id": "A00001",
  "bid_time": "2024-11-28T23:27:00Z",
  "bidder": {
    "id": "B00153",
    "rating": 83,
    "country": "AU"
  },
  "amount": 21.43,
  "device": {
    "os": "iOS",
    "browser": "Firefox"
  },
  "auto_bid_flag": false,
  "ip_address": "203.0.113.189"
}
```

Explain the structure of this record in words (what's at the top level vs. what's nested, and what fields each nested object contains).

**(b)**. Now let's load the data from the file `data/bids.jsonl` and examine the first two rows. Complete the starter code below and identify any nested columns:

::: {.callout-tip collapse = "true"}
## Streaming JSON has a different structure.

This JSON file has a different structure than we saw in the reading and in the lecture.
It's called streaming JSON (or sometimes "JSON Lines"): one complete JSON record per line, with no surrounding `[]` array and no commas between records.

Think of it like a log file: one complete JSON record per line.
(Regular JSON is often one big list: `[{...}, {...}, ...]`.)

Why it is used: it handles large files smoothly because you can read it line by line without loading everything.

How to load in R: use `stream_in()` (not `read_json()`).

Common gotcha: don't add brackets or commas between lines; each line must be valid JSON on its own.

:::

```{r}
#| eval: false
bids_json <- stream_in(file("YOUR_FILE_NAME"))

head(YOUR_DATA_SET, 2)
```

**(c)** Complete the code below to unnest the JSON data:

```{r}
#| eval: false
bids_json <-
    bids_json  |> 
    unnest_wider(YOUR_VARIABLE, names_sep = "_") |> 
    YOUR_FUNCTION(YOUR_VARIABLE, names_sep = "_") 

```

**(d)**. One possible review signal is multiple bidders using the same IP address. Explain why this could matter, then complete the code below to isolate those instances.

::: {.callout-tip collapse = "true"}
What's an IP address (and why we care)?

* Think of an IP address like a street address for a device on the internet. It helps websites know where to send data.

* On most shared Wi-Fi (homes, offices, libraries), many people/devices can appear under the same public IP. So “same IP” ≠ “same person.”

* IPs can change over time (dynamic addresses) and can be hidden/changed with VPNs — so they're imperfect identity signals.

:::

```{r}
#| eval: false
same_ip <-
    bids_json |>
    distinct(auction_id, ip_address, bidder_id) |>
    count(YOUR_VARIABLE, YOUR_VARIABLE) |>
    filter(YOUR_CONDITION) |>
    arrange(desc(n)) 
```

**(e)**. Another possible review signal is auto-bidding. Explain when auto-bids might be consistent with shill bidding versus legitimate behaviour. Complete the code below to measure the share of auto-bids per auction.

::: {.callout-tip collapse = true}
What is auto-bidding (proxy bidding)?

* It's a feature where you set a maximum price you're willing to pay and the system bids for you automatically in small steps to keep you on top—without you watching the auction.

* Why people use it: convenience and to avoid overbidding in the moment.

* What it looks like in the data: a series of quick, incremental bids appearing from the same bidder as others place bids.

Important: auto-bidding is normal. It's only review-worthy if you also see other flags (e.g., shared IPs, unusual timing patterns, repeated ties to the same seller).
:::

```{r}
#| eval: false
auto_bids <-
    bids_json |>
    group_by(auction_id) |>
    summarise(
        num_auto_bids = sum(YOUR_VARIABLE, na.rm = TRUE),
        total_bids    = YOUR_FUNCTION(),
        share_auto    = YOUR_VARIABLE / YOUR_VARIABLE
    ) |>
    arrange(desc(num_auto_bids))
```

**(f)** Let's combine the data on shared IP addresses and auto-bids into one data frame. Complete the code below, and then identify any auctions that might warrant review. Explain your answer.

```{r}
#| eval: false
# aggregate IP data to the auction level for easy join
same_ip_auctions <-
    same_ip |>
    group_by(auction_id) |>
    summarise(
        n_shared_ips = n(),           # how many IPs in this auction were used by 2+ bidders
        max_shared_n = max(n),        # largest number of bidders sharing a single IP
        .groups = "drop"
    )

shill_signals <- 
    same_ip_auctions |>
    left_join(YOUR_DATA, by = "auction_id") |>
    arrange(desc(n_shared_ips), desc(share_auto))

```


## In-Class Exercises

You will discuss these exercises in class with your peers in small groups and with your tutor. These exercises build from the exercises you have prepared above. You will get the most value from the class if you have completed those above before coming to class.

### Exercise 3: Opening the auction database

So far, we have worked with an auction-level CSV file and bid-level JSON Lines data. Those files are useful, but they each give us only one view of the platform.

The database gives us several linked tables: bids, auctions, items, users, and orders. This lets us ask relational questions, such as which bidder appears in which seller's auctions, without loading every table into memory at once. In this exercise, the goal is to orient yourself to the database and create a bidder-level activity summary that we will reuse in the next exercises.

::: {.callout-warning collapse=true}
## Quarto gotcha: close your DuckDB connection before you Render {.unnumbered}

Quarto renders your document in a fresh R session. If you opened a DuckDB connection in the Console (or earlier chunk) and leave it open, that session may lock the `.duckdb` file, causing errors like "database is locked" or "unable to open database file."

You can fix this by following one of these routes:

* Manage the connection inside the document and close it at the end

```{r}
#| eval: false
library(DBI)
library(duckdb)
con <- dbConnect(duckdb(), dbdir = "data/auctions.duckdb")

# ... your analysis ...

dbDisconnect(con, shutdown = TRUE)
```

If you connected in the Console, run:

```{r}
#| eval: false
dbDisconnect(con, shutdown = TRUE)
```

before clicking Render.

(If you still see a lock, especially on Windows, restart R and render again.)
:::

**(a)**. Explain how a database differs from the CSV and JSON data sources we have used so far. Why is a database useful for this part of the shill-bidding investigation?

**(b)**. Connect to the database stored in `data/auctions.duckdb` and list the tables present in it by completing the code below.

```{r}
#| eval: false
con <- dbConnect(duckdb(), dbdir = "YOUR_FILENAME")

dbListTables(YOUR_CODE)
```

**(c)**. Inspect the tables. For each table, write a short note describing what one row appears to represent and which key column could link it to another table.

```{r}
#| eval: false
tbl(con, "bids")       |> glimpse()
tbl(con, "YOUR_TABLE") |> glimpse()
tbl(con, "YOUR_TABLE") |> YOUR_FUNCTION()
YOUR_FUNCTION(con, "YOUR_TABLE") |> YOUR_FUNCTION()
YOUR_FUNCTION(YOUR_OBJECT, "YOUR_TABLE") |> YOUR_FUNCTION()
```

**(d)**. The first review question is simple: which bidders are most active? Complete the code below to count each bidder's total number of bids and the number of auctions they have participated in.

```{r}
#| eval: false
bidder_activity <- 
    tbl(con, "bids") |>
    group_by(YOUR_VARIABLE) |>
    YOUR_VARIABLE(
        total_bids = YOUR_FUNCTION(),
        auctions_participated = YOUR_FUNCTION(YOUR_VARIABLE),
        .groups = "drop"
    ) |>
    collect()
```

**(e)**. Inspect `bidder_activity`. Why is high activity useful as a starting point for review, and why is it not enough to suggest shill bidding on its own?


### Exercise 4: Measuring bidder-seller concentration

Activity tells us who bids a lot, but it does not tell us whether a bidder repeatedly supports the same seller. For shill-bidding review, that relationship matters: a bidder who focuses heavily on one seller's auctions may be more interesting than an equally active bidder who spreads bids across many sellers.

In this exercise, we will link bid events to sellers and compute each bidder's highest seller concentration: the share of that bidder's bids that go to their top seller.

**(a)**. A colleague suggests counting how often the same bidder bids against the same seller and across how many auctions. Why might these metrics be useful for identifying possible shill bidding behaviour? What innocent explanation could produce the same pattern?

**(b)**. We will instead compute the share of one bidder's bids that go to their top seller. What is the advantage of using a share rather than only raw bidder-seller counts?

**(c)**. Complete the code below. As you work through it, write one sentence explaining what each step contributes to the final `concentration` table.

```{r}
#| eval: false
# Step 1: rename to make easier to track
bids <- 
    tbl(con, "bids") |>
    rename(bidder_id = user_id)

# Step 2: link bids to the seller whose auction they are participating in
bids_items <- 
    bids |>
    left_join(tbl(con, "auctions") |> 
                  select(auction_id, item_id),
                by = "auction_id") |>
    left_join(tbl(con, "items")    |> 
                  select(item_id, seller_id),
                by = "item_id")

# Step 3: how many bids does each bidder make in total?
total_bids_per_bidder <- 
    bids_items |>
    YOUR_FUNCTION(bidder_id, name = "total_bids")

# Step 4: how many bids does each bidder make to each seller?
bidder_seller_bids <- 
    bids_items |>
    YOUR_FUNCTION(YOUR_VARIABLE, YOUR_VARIABLE, name = "bids_to_seller")

# Step 5: find each bidder's highest concentration on one seller
concentration <- 
    YOUR_DATASET |>
    YOUR_FUNCTION(YOUR_DATASET, 
              by = "bidder_id") |>
    mutate(seller_share = YOUR_VARIABLE / YOUR_VARIABLE) |>
    group_by(bidder_id) |> 
    slice_max(seller_share, n = 1) |> 
    collect()
```

**(d)**. Inspect `concentration`. Do any concentration ratios appear high enough to review? Identify a small set of bidders whose behaviour you would explore more closely, and explain your decision rule.

**(e)**. Why might a possible shill bidder's share of bids to one seller be an intermediate value, such as 20-30 percent, rather than above 50 percent?


### Exercise 5: Combining concentration with win rates

Concentration is useful, but it is still incomplete. A genuine buyer may focus on one seller because they like that seller's products. A possible shill bidder is more concerning when they appear often in a seller's auctions but rarely win. That pattern is consistent with pushing prices up while avoiding the final purchase.

In this exercise, we will compute bidder win rates, combine them with seller concentration, and use the result to identify review candidates.

**(a)**. Explain why win rate adds useful context to the concentration metric. Why should win rate be calculated using distinct auctions rather than raw bid rows?

**(b)**. Calculate how often bidders win the auctions they participate in.

```{r}
#| eval: false
# Step 1: winners per auction
winners <- 
    tbl(con, "orders") |>
    select(auction_id, winner_id)

# Step 2: distinct auctions each bidder joined
bidder_auctions <- 
    bids_items |>
    YOUR_FUNCTION(bidder_id, auction_id)

# Step 3: join winners and compute win_rate
win_rates <- 
    bidder_auctions |>
    YOUR_FUNCTION(winners, by = "auction_id") |>
    mutate(won = bidder_id == winner_id) |>
    group_by(bidder_id) |>
    summarise(
        auctions_participated = YOUR_FUNCTION(),
        auctions_won          = YOUR_FUNCTION(YOUR_VARIABLE, na.rm = TRUE),
        .groups = "drop"
    ) |>
    mutate(win_rate = YOUR_VARIABLE / YOUR_VARIABLE) |> 
    collect()
```

**(c)**. Join the `win_rates` data to the `concentration` data and save the combined dataset to `data/shill_metrics.csv`.

```{r}
#| eval: false
shill_metrics <-
    YOUR_DATASET |> 
    YOUR_FUNCTION(YOUR_DATASET, by = "bidder_id")

YOUR_FUNCTION(YOUR_DATASET, "data/shill_metrics.csv")
```

**(d)**. Create a scatter plot showing the relationship between `win_rate` and `seller_share`.

```{r}
#| eval: false
ggplot(shill_metrics) +
    geom_point(
        aes(x = YOUR_VARIABLE,
            y = YOUR_VARIABLE)
    ) +
    theme_minimal()
```

Which region of this plot would you treat as the strongest review zone for possible shill bidding? Explain why, and name one piece of additional evidence you would want before escalating the case.

**(e)**. Write a short manager-facing summary. Identify the type of bidder you would review first and explain why your language should describe the result as a review flag rather than proof of fraud.

**(f)** (Optional, if time). Propose an alternative way to use the `win_rate` and `concentration` data, along with any other data you have access to in the exercise, to identify possible shill bidders. Explain why your method works. You can explain your method intuitively; you do not need to code a solution.

When you finish the database work, close the DuckDB connection:

```{r}
#| eval: false
dbDisconnect(con, shutdown = TRUE)
```
