# Tutorial 9: Predictive Analytics for Workforce Planning {.unnumbered}

::: {.callout-tip}
## Learning Goals {.unnumbered}

By the end of this tutorial, you should be able to:

- Explore how attendance varies with game time (day/night) and opponent using tidy plots.
- Build time-based train/test splits and create cell-means predictors on the training set.
- Evaluate forecasts on the test set using MAE, RMSE, MAPE.
- Fit and compare regression models (Basic OLS, Quadratic in game number, Degree-8 polynomial) and relate fit vs. generalisation.
- Connect forecast accuracy to operational planning decisions.
:::

## The Business Challenge

```{r}
#| warning: false
#| echo: false
library(tidyverse)
library(slider)
library(ggokabeito)
```

```{r}
#| warning: false
#| eval: true
#| echo: false
#| cache: true

# Load the data
bos_2005_2015_fenway <- 
  read_csv('../data/mlb_attendance.csv')

```

### Using predictive analytics to navigate demand uncertainty and improve staffing decisions

In operations, managers must match capacity (people, time, equipment) to demand that arrives with uncertainty. Overstaffing raises costs; understaffing creates queues, slower service, and lost revenue. Prediction is the tool that turns historical data into forward-looking staffing guidance.

As in the lecture, our setting is MLB game attendance at Fenway Park, home of the Boston Red Sox. Concession managers need to staff bars and food stands for each Red Sox home game. The core uncertainty is attendance (how many fans show up). Attendance varies systematically with observable game features—e.g., day vs night, day of week, and opponent (rivalry effects)—and is bounded above by a capacity ceiling (sellouts cluster near the right edge of the distribution).

In this setting, the workflow that links demand forecasts to roster works as follows:

1. Forecast demand: Predict game-level attendance using historical patterns (e.g., cell-means or OLS models).
2. Translate to headcount: Convert predicted attendance into required staff using service rates (e.g., “one staff member can serve ~k customers per hour”).
3. Build the roster: Schedule staff to shifts ahead of time, subject to capacity and cost constraints.

Getting this workflow right is critical to business success. Accurate, well-calibrated forecasts let you set a baseline headcount and add a buffer sized to typical forecast error (e.g., MAE), striking a balance between cost and service quality.

In this week's workshop, we are going to get a feel for this workflow by tackling the following tasks:

- Exploration: Verify that day/night and opponent relate to attendance (so they’re worth using in models).
- Time-based splitting: Fit on earlier seasons, evaluate on later seasons to avoid look-ahead bias.
- Simple predictors: Use cell means (Day/Night; dotw × D/N; Opponent) for transparent, explainable baselines.
- Linear models: Compare Basic OLS, Quadratic in game number, and a Degree-8 polynomial—see how fit ≠ generalisation.
- Metrics: Judge models with MAE/RMSE/MAPE on the test set and connect those magnitudes to staffing buffers.

While doing so, we will keep our eye on the big picture: Our goal isn’t a perfect forecast; it’s a useful one—accurate enough, early enough, and explainable enough to make better staffing decisions.

Let's get started!

### Getting set up

First, we need to load our packages:

```{r}
#| eval: false

library(tidyverse)
library(scales)
library(ggokabeito)
```

Next, we need to load our dataset.

```{r}
#| eval: false

bos_2005_2015_fenway <- 
  read_csv('data/mlb_attendance.csv')

```

## Prepare these Exercises before Class

Prepare these exercises before coming to class. Plan to spend 30 minutes on these exercises.

### Exercise 1: Building advance predictors from the training set

In the lecture, the lag-1 model performed well because recent attendance is informative. But it is not always useful for workforce planning: concession managers often need to roster staff before the previous game has been played. In this tutorial we therefore focus on predictors that are known before the game:

- whether the game is during the day or evening
- the day of the week
- the opponent

Our goal is to use the training data to decide whether these forward-looking features contain useful predictive signal, then turn them into simple test-set forecasts.

::: {.callout-note title="Why split into training and test data?" collapse="true"}
**Overfitting** happens when a model learns patterns that are too specific to the data used to build it. In practice, this means the model is fitting idiosyncratic variation in the training data rather than meaningful structure. Idiosyncratic variation is not systematic, so we should not expect it to reoccur in future games in the same way. The model may therefore look accurate on familiar data, but perform poorly when asked to predict new games.

A training/test split helps us check for this. We build the model using the **training set**, then evaluate it on a separate **test set** that was not used to calculate the model. If the model works well on the test set, we have better evidence that it has learned a useful pattern rather than memorised noise.

Because this is a forecasting problem, we split by time: earlier seasons are used for training, and later seasons are held out for testing. This mirrors how managers would actually use the model: learn from the past, then forecast future games.
:::

**(a).** Create a time-based training and test split. Use games from seasons before 2013 as the training set, and games from 2013 onward as the test set.

```{r}
#| eval: false

training_sample_2005_2012 <- 
  bos_2005_2015_fenway |>
  YOUR_CODE_HERE  

test_sample_2013_2015 <- 
  bos_2005_2015_fenway |>
  YOUR_CODE_HERE  
```

Why is a time-based split more realistic than randomly splitting games for this forecasting problem?

**(b).** Start with the simplest forward-looking feature: whether the game is played during the day or at night. Complete and run the code below, then describe the size and direction of the difference in attendance across day vs night games.

```{r}
#| eval: false

training_sample_2005_2012 |>
  YOUR_CODE_HERE(day_night) |>
  YOUR_CODE_HERE(
    ave_attendance = mean(attendance, na.rm = TRUE), 
    n_games = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = day_night, y = ave_attendance, fill = day_night)) +
  geom_col(width = 0.5) +
  geom_text(
    aes(label = round(ave_attendance, 0)),
    vjust = -0.5, size = 4, color = "black"
  ) +
  labs(x = "Game Time", y = "Average Attendance", fill = "Game Time") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  ggokabeito::scale_fill_okabe_ito() +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

**(c).** Now inspect two richer versions of the same idea: attendance across the different day of week-game time combinations, and attendance across the different opponents. Run the code below.

```{r}
#| eval: false

training_sample_2005_2012 |>
  mutate(
    dotw = factor(
      dotw,
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    )
  ) |>
  group_by(dotw, day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = dotw, y = ave_attendance, fill = day_night)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  ggokabeito::scale_fill_okabe_ito(name = "Game Time") +
  labs(x = "Day of the Week", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    n_games = n(),
    .groups = "drop"
  ) |>
  arrange(desc(ave_attendance)) |>
  mutate(opp = factor(opp, levels = opp)) |>
  ggplot(aes(x = opp, y = ave_attendance)) +
  geom_col(fill = "steelblue", width = 0.5) +
  labs(x = "Opponent", y = "Average Attendance") +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 5)) +
  coord_cartesian(ylim = c(30000, 40000)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1)
  )
```

Briefly answer the following:

- Which feature appears most useful for forecasting attendance?
- Which feature would be easiest to explain to a concessions manager?
- Which feature is most likely to be noisy because it creates small groups?

**(d).** Build three cell-means models from the training set, then join those training means to the test set as predictions. 

::: {.callout-note title="What is a cell-means model?" collapse="true"}
A cell-means model predicts using the average outcome within a group, or "cell", in the training data. For example, a day/night cell-means model calculates the average training-set attendance for day games and the average training-set attendance for night games. It then uses those averages as predictions for future day and night games.

This kind of model is simple and explainable: "games like this usually draw about this many fans." The tradeoff is that more detailed (i.e., narrower) cells can become noisy.
:::

Complete the missing pieces in the code below to fit these cell-means models.

```{r}
#| eval: false

average_attendance_by_daynight <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_dotw_dn <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE, YOUR_CODE_HERE) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_opp <- 
  training_sample_2005_2012 |>
  group_by(YOUR_CODE_HERE) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_daynight, by = YOUR_CODE_HERE) |>
  rename(predicted_attendance_daynight = ave_attendance) |>
  left_join(average_attendance_by_dotw_dn, by = c(YOUR_CODE_HERE, YOUR_CODE_HERE)) |>
  rename(predicted_attendance_dotw_dn = ave_attendance) |>
  left_join(average_attendance_by_opp, by = YOUR_CODE_HERE) |>
  rename(predicted_attendance_opp = ave_attendance)
```

**(e).** Each model uses information available before the game. Before calculating formal error metrics, which one do you expect to perform best on the test set, and why? In your answer, distinguish between a feature that looks useful in the training data and a model that is likely to generalise well to future games.

```{r}
# Write your answer here
```

### Exercise 2: Evaluating advance predictors in the test set

The lecture introduced MAE, RMSE, and MAPE. Rather than rebuild the same error pipeline three separate times, we will use one tidy workflow to compare all three advance-predictor models from Exercise 1.

**(a).** Run the code below to put the three prediction columns into a long format. Before doing so complete the missing lines that calculate forecast errors.

```{r}
#| eval: false

simple_model_errors <- 
  test_sample_2013_2015 |>
  select(
    attendance,
    predicted_attendance_daynight,
    predicted_attendance_dotw_dn,
    predicted_attendance_opp
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_daynight = "Day/Night",
      predicted_attendance_dotw_dn = "Day of week x Day/Night",
      predicted_attendance_opp = "Opponent"
    ),
    error            = YOUR_CODE_HERE - YOUR_CODE_HERE,
    abs_error        = abs(YOUR_CODE_HERE),
    squared_error    = YOUR_CODE_HERE ^ 2,
    percentage_error = (YOUR_CODE_HERE / YOUR_CODE_HERE) * 100
  )
```

What does `pivot_longer()` do here, and why is it helpful when we want to compare several models?

**(b).** Complete and run the code below to calculate MAE, RMSE, and MAPE for each model.

```{r}
#| eval: false

simple_model_metrics <- 
  simple_model_errors |>
  group_by(model) |>
  summarise(
    MAE  = mean(YOUR_CODE_HERE, na.rm = TRUE),
    RMSE = sqrt(mean(YOUR_CODE_HERE, na.rm = TRUE)),
    MAPE = mean(abs(YOUR_CODE_HERE), na.rm = TRUE),
    .groups = "drop"
  )
```

Which metric is most directly useful for setting a staffing buffer? Which metric is most sensitive to occasional large misses?

**(c).** Run the code below to plot the metrics.

```{r}
#| eval: false

simple_metrics_long <-
  simple_model_metrics |>
  pivot_longer(
    cols = c(MAE, RMSE, MAPE),
    names_to = "metric",
    values_to = "value"
  )

ggplot(simple_metrics_long, aes(x = model, y = value, fill = model)) +
  geom_col() +
  geom_text(aes(label = round(value, 0)), vjust = -0.5, size = 3) +
  facet_wrap(~ metric, scales = "free_y") +
  ggokabeito::scale_fill_okabe_ito() +
  labs(
    title = "Forecast Accuracy Metrics by Model",
    x = "Model",
    y = "Value"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**(d).** Based on the table and plot, choose one model for a concessions staffing manager. In your answer, consider:

- test-set accuracy
- whether the predictor is available early enough for rostering
- whether the model is simple enough to explain
- whether the accuracy differences are large enough to matter operationally

```{r}
# Write your answer here
```

## In-Class Exercises

You will discuss these exercises in class with your peers in small groups and with your tutor. These exercises build from the exercises you have prepared above, you will get the most value from the class if you have completed those above before coming to class.

### Exercise 3: Fitting regression models

Exercises 1 and 2 used simple cell-means models: each forecast came from the average attendance for one group of games. We now turn to ordinary least squares (OLS) regression, which among other things lets us combine multiple predictors in a single model.

The aim is not to find the most complicated model possible. The aim is to ask whether a richer model improves the forecast enough to justify the extra complexity for staffing decisions.

::: {.callout-note title="What is OLS regression?" collapse="true"}
Ordinary least squares (OLS) regression is a way to model the relationship between an outcome we want to predict and one or more predictors. Here, the outcome is attendance, and the predictors might include opponent, game time, day of week, or game number.

OLS works by choosing the line, curve, or set of group differences that makes the model's prediction errors as small as possible in the training data. More precisely, it chooses the coefficients that minimise the sum of squared errors: the squared differences between the actual attendance values and the model's fitted values.

For example, a simple regression might estimate how attendance changes between day and night games. A richer regression can combine several predictors at once, such as opponent, day/night, and day of week. When a predictor is categorical, such as opponent, OLS compares each category to a reference category.

In this tutorial, we use OLS as a forecasting tool. We fit the model on earlier seasons, use it to predict attendance in later seasons, and then evaluate whether those predictions are accurate enough and explainable enough to support staffing decisions.
:::

**(a).** Fit an OLS model that uses the same forward-available predictors we studied earlier: opponent, day/night, and day of week. Complete and run the code below.

```{r}
#| eval: false

basic_ols_model <- 
  lm(YOUR_CODE_HERE ~ YOUR_CODE_HERE + YOUR_CODE_HERE + YOUR_CODE_HERE, 
     data = YOUR_CODE_HERE                                               
  )
```

**(b).** Run the code below to view the model summary.

```{r}
#| eval: false

summary(basic_ols_model)
```

Discuss with your group:

- Which predictors appear to have the largest associations with attendance?
- What does the R-squared value tell us about in-sample fit?
- Why should we avoid choosing the model on R-squared alone?

**(c).** The basic OLS model combines categorical predictors. We can also ask whether attendance changes systematically over the course of the season. Run the two plots below to compare a simple quadratic curve with a more flexible degree-8 curve.

```{r}
#| eval: false

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x^2),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Quadratic Fit of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()

ggplot(training_sample_2005_2012, aes(x = gm, y = attendance)) +
  geom_point(alpha = 0.6) +
  stat_smooth(
    method  = "lm",
    formula = y ~ x + I(x ^ 2) + I(x ^ 3) + I(x ^ 4) + 
      I(x ^ 5) + I(x ^ 6) + I(x ^ 7) + I(x ^ 8),
    color   = "blue",
    se      = FALSE
  ) +
  labs(
    title = "Degree-8 Polynomial of Attendance by Game Number",
    x     = "Game Number",
    y     = "Attendance"
  ) +
  theme_minimal()
```

Which curve looks more likely to generalise to future games, and why?

**(d).** Let's formally model the plots above. Run the code below to fit two additional OLS specifications: a simple quadratic model using game number, and a more flexible specification that combines opponent, day/night, day of week, and a degree-8 polynomial in game number.

```{r}
#| eval: false

non_linear_ols_model <- 
  lm(attendance ~ gm + I(gm^2), 
     data = training_sample_2005_2012
  )

eighth_poly_ols_model <- 
  lm(attendance ~ opp + day_night + dotw + gm + I(gm^2) + 
     I(gm^3) + I(gm^4) + I(gm^5) + I(gm^6) + I(gm^7) + I(gm^8),
     data = training_sample_2005_2012
  )
```

**(e).** Complete and run the code below to compare in-sample R-squared across the three OLS specifications.

```{r}
#| eval: false

ols_fit_summary <- tibble(
  model = c("Basic OLS", "Quadratic game number", "Degree-8 polynomial OLS"),
  r_squared = c(
    summary(YOUR_CODE_HERE)$r.squared,
    summary(YOUR_CODE_HERE)$r.squared,
    summary(YOUR_CODE_HERE)$r.squared
  )
)

ols_fit_summary
```

Which model fits the training data best? Why is this not enough to choose the forecasting model? In your answer, consider fit, complexity, and operational usefulness.

```{r}
# Write your answer here
```

### Exercise 4: Evaluating regression models

Now that we have fitted our OLS models, we need to test whether the extra complexity actually improves forecasts for future games. This mirrors Exercise 2: train on earlier seasons, evaluate on later seasons, and choose based on usefulness for staffing rather than in-sample fit alone.

**(a).** Complete and run the setup code below. It creates a test sample that contains only opponent levels observed in the training data, then generates test-set predictions from the three OLS models.

```{r}
#| eval: false

known_opp_levels <- 
  training_sample_2005_2012 |> 
  pull(opp) |> 
  unique()

safe_test_sample <- 
  test_sample_2013_2015 |>
  filter(opp %in% known_opp_levels)

safe_test_sample <-
  safe_test_sample |>
  mutate(
    predicted_attendance_basic_OLS =
      predict(YOUR_CODE_HERE, newdata = safe_test_sample),
    predicted_attendance_non_linear_OLS =
      predict(YOUR_CODE_HERE, newdata = safe_test_sample),
    predicted_attendance_eighth_poly_OLS =
      predict(YOUR_CODE_HERE, newdata = safe_test_sample)
  )
```

Why do we need to handle opponent levels carefully before using `predict()` on the test set? What would happen if we didn't first trim our test sample in this way?

**(b).** Run the code below, then inspect the resulting `ols_model_errors` data frame.

```{r}
#| eval: false

ols_model_errors <- 
  safe_test_sample |>
  select(
    attendance,
    predicted_attendance_basic_OLS,
    predicted_attendance_non_linear_OLS,
    predicted_attendance_eighth_poly_OLS
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_basic_OLS = "Basic OLS",
      predicted_attendance_non_linear_OLS = "Quadratic game number",
      predicted_attendance_eighth_poly_OLS = "Degree-8 polynomial OLS"
    ),
    error            = attendance - predicted_attendance,
    abs_error        = abs(error),
    squared_error    = error ^ 2,
    percentage_error = (error / attendance) * 100
  )
```

What does this code do? Why is each row now a model-game combination rather than just a game? Why do we create `abs_error`, `squared_error`, and `percentage_error`?

**(c).** Now use `ols_model_errors` to create `ols_model_metrics`, a table like the one below. Your output should show the values for each cell - e.g., each model's MAE, RMSE, etc.

| model | MAE | RMSE | MAPE |
|---|---:|---:|---:|
| Quadratic game number |  |  |  |
| Degree-8 polynomial OLS |  |  |  |
| Basic OLS |  |  |  |

Write the code to produce this table:

```{r}
#| eval: false

# Write your answer here

```

**(d).** Interpret the table from part (c).

Which OLS model performs best out of sample? Why doesn't this match the ranking by training-set R-squared from Exercise 3?

```{r}
# Write your answer here
```

### Exercise 5: Choosing and using a staffing forecast

The final decision is not just "which OLS model won?" We also have the simple cell-means models from Exercise 2. Before comparing them, we need to check that the models were evaluated on the same games.

**(a).** In Exercise 2, the simple cell-means models were evaluated on `test_sample_2013_2015`. In Exercise 4, the OLS models were evaluated on `safe_test_sample`.

Run the code below to compare the two samples.

```{r}
#| eval: false

tibble(
  sample = c("Full test sample", "OLS-safe test sample"),
  n_games = c(
    nrow(test_sample_2013_2015),
    nrow(safe_test_sample)
  )
)

test_sample_2013_2015 |>
  filter(!(opp %in% known_opp_levels)) |>
  count(opp, sort = TRUE)
```

What is the sample mismatch?

**(b).** The table below fixes the sample mismatch from part (a). It compares the best simple cell-means model with the best OLS model, using only the same OLS-safe test rows.

```{r}
#| echo: false
#| message: false
#| warning: false

training_sample_2005_2012 <- 
  bos_2005_2015_fenway |>
  filter(year < 2013)

test_sample_2013_2015 <- 
  bos_2005_2015_fenway |>
  filter(year >= 2013)

average_attendance_by_daynight <- 
  training_sample_2005_2012 |>
  group_by(day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_dotw_dn <- 
  training_sample_2005_2012 |>
  group_by(dotw, day_night) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

average_attendance_by_opp <- 
  training_sample_2005_2012 |>
  group_by(opp) |>
  summarise(
    ave_attendance = mean(attendance, na.rm = TRUE),
    .groups = "drop"
  )

test_sample_2013_2015 <- 
  test_sample_2013_2015 |>
  left_join(average_attendance_by_daynight, by = "day_night") |>
  rename(predicted_attendance_daynight = ave_attendance) |>
  left_join(average_attendance_by_dotw_dn, by = c("dotw", "day_night")) |>
  rename(predicted_attendance_dotw_dn = ave_attendance) |>
  left_join(average_attendance_by_opp, by = "opp") |>
  rename(predicted_attendance_opp = ave_attendance)

basic_ols_model <- 
  lm(attendance ~ opp + day_night + dotw,
     data = training_sample_2005_2012
  )

non_linear_ols_model <- 
  lm(attendance ~ gm + I(gm^2), 
     data = training_sample_2005_2012
  )

eighth_poly_ols_model <- 
  lm(attendance ~ opp + day_night + dotw + gm + I(gm^2) + 
     I(gm^3) + I(gm^4) + I(gm^5) + I(gm^6) + I(gm^7) + I(gm^8),
     data = training_sample_2005_2012
  )

known_opp_levels <- 
  training_sample_2005_2012 |> 
  pull(opp) |> 
  unique()

safe_test_sample <- 
  test_sample_2013_2015 |>
  filter(opp %in% known_opp_levels)

safe_test_sample <- 
  safe_test_sample |>
  mutate(
    predicted_attendance_basic_OLS =
      predict(basic_ols_model, newdata = safe_test_sample),
    predicted_attendance_non_linear_OLS =
      predict(non_linear_ols_model, newdata = safe_test_sample),
    predicted_attendance_eighth_poly_OLS =
      predict(eighth_poly_ols_model, newdata = safe_test_sample)
  )

comparable_simple_model_metrics <- 
  safe_test_sample |>
  select(
    attendance,
    predicted_attendance_daynight,
    predicted_attendance_dotw_dn,
    predicted_attendance_opp
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_daynight = "Day/Night",
      predicted_attendance_dotw_dn = "Day of week x Day/Night",
      predicted_attendance_opp = "Opponent"
    ),
    error            = attendance - predicted_attendance,
    abs_error        = abs(error),
    squared_error    = error ^ 2,
    percentage_error = (error / attendance) * 100
  ) |>
  group_by(model) |>
  summarise(
    MAE  = mean(abs_error, na.rm = TRUE),
    RMSE = sqrt(mean(squared_error, na.rm = TRUE)),
    MAPE = mean(abs(percentage_error), na.rm = TRUE),
    .groups = "drop"
  )

ols_model_metrics <- 
  safe_test_sample |>
  select(
    attendance,
    predicted_attendance_basic_OLS,
    predicted_attendance_non_linear_OLS,
    predicted_attendance_eighth_poly_OLS
  ) |>
  pivot_longer(
    cols = starts_with("predicted_attendance_"),
    names_to = "model",
    values_to = "predicted_attendance"
  ) |>
  mutate(
    model = recode(
      model,
      predicted_attendance_basic_OLS = "Basic OLS",
      predicted_attendance_non_linear_OLS = "Quadratic game number",
      predicted_attendance_eighth_poly_OLS = "Degree-8 polynomial OLS"
    ),
    error            = attendance - predicted_attendance,
    abs_error        = abs(error),
    squared_error    = error ^ 2,
    percentage_error = (error / attendance) * 100
  ) |>
  group_by(model) |>
  summarise(
    MAE  = mean(abs_error, na.rm = TRUE),
    RMSE = sqrt(mean(squared_error, na.rm = TRUE)),
    MAPE = mean(abs(percentage_error), na.rm = TRUE),
    .groups = "drop"
  )

best_simple_model <-
  comparable_simple_model_metrics |>
  arrange(MAE) |>
  slice(1) |>
  mutate(model_family = "Cell-means")

best_ols_model <-
  ols_model_metrics |>
  arrange(MAE) |>
  slice(1) |>
  mutate(model_family = "OLS")

best_model_comparison <-
  best_simple_model |>
  bind_rows(best_ols_model) |>
  select(model_family, model, MAE, RMSE, MAPE)

best_model_comparison_display <-
  best_model_comparison |>
  transmute(
    `Model family` = model_family,
    `Best model in family` = model,
    `MAE (fans)` = format(round(MAE, 1), nsmall = 1),
    `RMSE (fans)` = format(round(RMSE, 1), nsmall = 1),
    `MAPE (%)` = format(round(MAPE, 1), nsmall = 1)
  )

knitr::kable(
  best_model_comparison_display,
  caption = "Best cell-means model versus best OLS model on the same OLS-safe test rows"
)
```

What does this table compare? Why is this comparison fair, while a direct comparison between `simple_model_metrics` from Exercise 2 and `ols_model_metrics` from Exercise 4 was not fair?

**(c).** Interpret the table in part (b).

Which model has the lower MAE on the same test rows? Is the difference large enough to matter for a concessions staffing decision? Would RMSE or MAPE change your conclusion?

**(d).** Recommend a model and explain how the manager should use it.

In your answer, include:

- which model you choose
- the evidence from the same-sample metrics
- why the extra complexity is or is not worth it
- how the forecast becomes a baseline roster
- how MAE would be used as a staffing buffer and monitored over time

**(e).** A good working model still needs maintenance. Explain when and why the manager should update or replace the model.

In your answer, consider:

- what the manager should monitor over time
- what changes might make the model stale
- how a major change in the environment could affect forecast accuracy

```{r}
# Write your answer here
```
